---
title: "Mississippi Enrollment Trends"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mississippi Enrollment Trends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

```{r load-packages}
library(msschooldata)
library(ggplot2)
library(dplyr)
library(scales)
```

```{r theme}
theme_readme <- function() {
  theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "gray40"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
}

colors <- c("total" = "#2C3E50", "white" = "#3498DB", "black" = "#E74C3C",
            "hispanic" = "#F39C12", "asian" = "#9B59B6")
```

```{r fetch-data, cache = TRUE}
# Get available years
years <- get_available_years()
if (is.list(years)) {
  max_year <- years$max_year
  min_year <- years$min_year
} else {
  max_year <- max(years)
  min_year <- min(years)
}

# Fetch data with error handling for CI environments
enr <- tryCatch({
  fetch_enr_multi((max_year - 9, use_cache = TRUE):max_year)
}, error = function(e) {
  message("Note: Could not fetch enrollment data - ", e$message)
  NULL
})

key_years <- seq(max(min_year, 2007), max_year, by = 5)
if (!max_year %in% key_years) key_years <- c(key_years, max_year)

enr_long <- tryCatch({
  fetch_enr_multi(key_years, use_cache = TRUE)
}, error = function(e) NULL)

enr_current <- tryCatch({
  fetch_enr(max_year, use_cache = TRUE)
}, error = function(e) NULL)

# Check if we have data
has_data <- !is.null(enr) && nrow(enr) > 0
```

## 1. Mississippi is majority Black in many districts

Unlike most Southern states, Mississippi has numerous majority-Black school districts, especially in the Delta region.

```{r majority-black, eval=has_data}
black <- enr_current %>%
  filter(is_district, subgroup == "black", grade_level == "TOTAL") %>%
  arrange(desc(pct)) %>%
  head(10) %>%
  mutate(district_label = reorder(district_name, pct))

ggplot(black, aes(x = district_label, y = pct * 100)) +
  geom_col(fill = colors["black"]) +
  coord_flip() +
  labs(title = "Mississippi Has Many Majority-Black Districts",
       subtitle = "Especially in the Delta region",
       x = "", y = "Percent Black Students") +
  theme_readme()
```

## 2. The Delta is emptying out

Districts in the Mississippi Delta have lost 30-50% of students since 2007.

```{r delta-decline, eval=has_data}
delta <- c("Coahoma County", "Bolivar County", "Sunflower County", "Leflore County")
delta_trend <- enr_long %>%
  filter(is_district, grepl(paste(delta, collapse = "|"), district_name, ignore.case = TRUE),
         subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  group_by(end_year) %>%
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")

ggplot(delta_trend, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma) +
  labs(title = "The Delta is Emptying Out",
       subtitle = "Coahoma, Bolivar, Sunflower, Leflore counties combined",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 3. DeSoto County: Mississippi's growth engine

Bordering Memphis, DeSoto County has nearly doubled enrollment.

```{r desoto-growth, eval=has_data}
desoto <- enr %>%
  filter(is_district, grepl("DeSoto", district_name, ignore.case = TRUE),
         subgroup == "total_enrollment", grade_level == "TOTAL")

ggplot(desoto, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "DeSoto County: Mississippi's Growth Engine",
       subtitle = "Memphis suburb nearly doubled enrollment",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 4. Jackson Public Schools' steep decline

Mississippi's capital city has lost over 40% of students.

```{r jackson-decline, eval=has_data}
jackson <- enr %>%
  filter(is_district, grepl("Jackson Public", district_name, ignore.case = TRUE),
         subgroup == "total_enrollment", grade_level == "TOTAL")

ggplot(jackson, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "Jackson Public Schools' Steep Decline",
       subtitle = "Capital city lost over 40% of students",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 5. Economic disadvantage is nearly universal

Over 75% of Mississippi students are economically disadvantaged - the highest rate in the nation.

```{r econ-disadvantage, eval=has_data}
econ <- enr %>%
  filter(is_state, subgroup == "econ_disadv", grade_level == "TOTAL")

ggplot(econ, aes(x = end_year, y = pct * 100)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  labs(title = "Economic Disadvantage is Nearly Universal",
       subtitle = "Over 75% of Mississippi students - highest in the nation",
       x = "School Year", y = "Percent Economically Disadvantaged") +
  theme_readme()
```

## 6. COVID hit kindergarten hard

Mississippi lost 7% of kindergartners in 2021 and enrollment hasn't recovered.

```{r covid-k, eval=has_data}
k_trend <- enr %>%
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% c("K", "01", "06", "12")) %>%
  mutate(grade_label = case_when(
    grade_level == "K" ~ "Kindergarten",
    grade_level == "01" ~ "Grade 1",
    grade_level == "06" ~ "Grade 6",
    grade_level == "12" ~ "Grade 12"
  ))

ggplot(k_trend, aes(x = end_year, y = n_students, color = grade_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  geom_vline(xintercept = 2021, linetype = "dashed", color = "red", alpha = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "COVID Hit Mississippi Kindergarten Hard",
       subtitle = "Lost 7% of kindergartners and hasn't recovered",
       x = "School Year", y = "Students", color = "") +
  theme_readme()
```

## 7. Madison County: Suburban success

Madison County has grown while Jackson shrinks - classic suburban flight.

```{r madison-growth, eval=has_data}
madison <- enr %>%
  filter(is_district, grepl("Madison County", district_name, ignore.case = TRUE),
         subgroup == "total_enrollment", grade_level == "TOTAL")

ggplot(madison, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "Madison County: Suburban Success",
       subtitle = "Growing while Jackson shrinks - classic suburban flight",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 8. Hispanic population is growing

From 2% to over 4% statewide, with some districts like Forest Municipal reaching 20%+.

```{r hispanic-growth, eval=has_data}
hispanic <- enr_current %>%
  filter(is_district, subgroup == "hispanic", grade_level == "TOTAL") %>%
  arrange(desc(pct)) %>%
  head(10) %>%
  mutate(district_label = reorder(district_name, pct))

ggplot(hispanic, aes(x = district_label, y = pct * 100)) +
  geom_col(fill = colors["hispanic"]) +
  coord_flip() +
  labs(title = "Hispanic Population Growing",
       subtitle = "Some districts like Forest Municipal reaching 20%+",
       x = "", y = "Percent Hispanic Students") +
  theme_readme()
```

## 9. The Coast is holding steady

Gulf Coast districts have maintained enrollment despite hurricanes.

```{r coast-stable, eval=has_data}
coast <- c("Harrison County", "Jackson County", "Hancock County")
coast_trend <- enr %>%
  filter(is_district, grepl(paste(coast, collapse = "|"), district_name, ignore.case = TRUE),
         subgroup == "total_enrollment", grade_level == "TOTAL")

ggplot(coast_trend, aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "The Coast is Holding Steady",
       subtitle = "Gulf Coast districts maintained enrollment despite hurricanes",
       x = "School Year", y = "Students", color = "") +
  theme_readme()
```

## 10. Charter schools are minimal

Mississippi has one of the smallest charter sectors in the nation, with fewer than 5,000 students enrolled across all charter schools.

*Note: Charter school enrollment tracking is not yet implemented in this package. The MDE data portal does not currently distinguish charter schools as a separate entity type.*
